New Metrics for Assessing Projection Pursuit Indexes, and Guiding Optimisation Choices

### 2024-09-11 jso.tex
Table 1 math notations
GP parameters' names replaced by their mathematical notations in Table 1

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{orcidlink}
\usepackage{algorithm}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\theoremstyle{plain}
\newtheorem{assumption}{\protect\assumptionname}
\newtheorem{claim}{\protect\claimname}
\newtheorem{condition}{\protect\conditionname}
\newtheorem{conjecture}{\protect\conjecturename}
\newtheorem{cor}{\protect\corollaryname}
\newtheorem{defn}{\protect\definitionname}
\newtheorem{example}{\protect\examplename}
\newtheorem{lem}{\protect\lemmaname}
\newtheorem{notation}{\protect\notationname}
\newtheorem{problem}{\protect\problemname}
\newtheorem{prop}{\protect\propositionname}
\newtheorem{rem}{\protect\remarkname}
\newtheorem{thm}{\protect\theoremname}
\providecommand{\assumptionname}{Assumption}
\providecommand{\claimname}{Claim}
\providecommand{\conditionname}{Condition}
\providecommand{\conjecturename}{Conjecture}
\providecommand{\corollaryname}{Corollary}
\providecommand{\definitionname}{Definition}
\providecommand{\examplename}{Example}
\providecommand{\lemmaname}{Lemma}
\providecommand{\notationname}{Notation}
\providecommand{\problemname}{Problem}
\providecommand{\propositionname}{Proposition}
\providecommand{\remarkname}{Remark}
\providecommand{\theoremname}{Theorem}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={New Metrics for Assessing Projection Pursuit Indexes, and Guiding Optimisation Choices},
  pdfauthor={H. Sherry Zhang; Dianne Cook; Nicolas Langrené; Jessica Wai Yin Leung},
  pdfkeywords={projection pursuit, jellyfish search optimiser
(JSO), optimisation, grand tour, high-dimensional data, exploratory data
analysis},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@austin.utexas.edu}{\nolinkurl{huize.zhang@austin.utexas.edu}}. Dianne
Cook. Email: \href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}. Nicolas
Langrené. Email: \href{mailto:nicolaslangrene@uic.edu.cn}{\nolinkurl{nicolaslangrene@uic.edu.cn}}. Jessica
Wai Yin
Leung. Email: \href{mailto:Jessica.Leung@monash.edu}{\nolinkurl{Jessica.Leung@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Statistics and Data
Sciences, University of Texas at Austin, Austin, United
States\\ \textsuperscript{2} Department of Econometrics and Business
Statistics, Monash
University, Melbourne, Australia\\ \textsuperscript{3} Department of
Mathematical Sciences, Guangdong Provincial Key Laboratory of
Interdisciplinary Research and Application for Data Science,
BNU-HKBU~United~International~College, Zhuhai, China
\begin{abstract}
The projection pursuit (PP) guided tour interactively optimises a
criteria function known as the PP index, to explore high-dimensional
data by revealing interesting projections. Optimisation of some PP
indexes can be non-trivial, if they are non-smooth functions, or the
optima has a small ``squint angle'', detectable only from close
proximity. To address these challenges, this study investigates the
performance of a recently introduced swarm-based algorithm, Jellyfish
Search Optimiser (JSO), for optimising PP indexes. The performance of
JSO for visualising data is evaluated across various hyper-parameter
settings and compared with existing optimisers. Additionally, methods
for calculating the smoothness and squintability properties of the PP
index are proposed. They are used to assess the optimiser performance in
the presence of PP index complexities. A simulation study illustrates
the use of these performance metrics to compare the JSO with existing
optimisation methods available for the guided tour. The JSO algorithm
has been implemented in the R package, \texttt{tourr}, and functions to
calculate smoothness and squintability are available in the
\texttt{ferrn} package.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
projection pursuit\sep jellyfish search optimiser
(JSO)\sep optimisation\sep grand tour\sep high-dimensional data\sep 
exploratory data analysis
\end{keywords}

\setstretch{2}
\section{Introduction}\label{introduction}

@@

\section{Properties of PP indexes}\label{sec-PP-properties}

@@

\section{The jellyfish optimiser}\label{sec-JSO}

@@

\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, colback=white, rightrule=.15mm, opacitybacktitle=0.6, leftrule=.75mm, coltitle=black, breakable, title={Algorithm: Jellyfish Optimizer Pseudo Code}, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, opacityback=0, left=2mm, bottomtitle=1mm, titlerule=0mm, arc=.35mm, toprule=.15mm, bottomrule=.15mm]

@@

\begin{table}

\caption{\label{tbl-smoothness-squintability}Parameters estimated from
the Gaussian process (including variance, range, smoothness, and nugget)
and scaled logistic function (\(\theta_1\) to \(\theta_4\)) for the
pipe-finding and sine-wave finding problems. The column ``smooth'' and
``squint'' represent the smoothness and squintability measures.}

\centering{

@@

\begin{tabular}{|>{}lll>{}r|rrr>{}r|rrrr>{}r|}
\toprule
  & shape & index & d & variance & range & smooth & nugget & $\theta_1$ & $\theta_2$ & $\theta_3$ & $\theta_4$ & squint\\
\midrule
1 & pipe & holes & 6 & 0.00 & 0.37 & 2.36 & 0.22 & 1.02 & 0.86 & 3.37 & 0.02 & 3.05\\
2 & pipe & holes & 8 & 0.00 & 0.18 & 2.19 & 0.82 & 1.01 & 0.87 & 3.26 & 0.03 & 2.96\\
3 & pipe & holes & 10 & 0.00 & 0.11 & 2.19 & 3.03 & 1.02 & 0.88 & 3.15 & 0.02 & 2.95\\
4 & pipe & holes & 12 & 0.00 & 0.15 & 2.29 & 1.58 & 1.01 & 0.88 & 3.34 & 0.00 & 3.12\\
5 & sine & MIC & 6 & 0.02 & 0.09 & 2.46 & 0.08 & 0.89 & 0.57 & 1.62 & -0.02 & 1.26\\
6 & sine & MIC & 8 & 0.02 & 0.08 & 2.64 & 0.08 & 0.93 & 0.33 & 1.31 & -0.03 & 0.77\\
7 & sine & TIC & 6 & 0.12 & 0.11 & 2.44 & 0.09 & 0.95 & 0.54 & 1.72 & -0.03 & 1.32\\
8 & sine & TIC & 8 & 0.12 & 0.10 & 2.47 & 0.09 & 0.95 & 0.56 & 1.72 & -0.03 & 1.37\\
9 & sine & dcor2d & 6 & 0.03 & 0.17 & 2.66 & 0.11 & 0.95 & 1.04 & 2.74 & -0.02 & 2.95\\
10 & sine & loess2d & 6 & 0.08 & 0.34 & 1.99 & 0.31 & 1.02 & 1.04 & 2.65 & 0.08 & 2.76\\
11 & sine & splines2d & 6 & 0.04 & 0.24 & 2.54 & 0.10 & 1.01 & 1.05 & 2.73 & -0.01 & 3.12\\
12 & sine & stringy & 6 & 0.00 & 0.01 & 1.54 & 38.17 & 1.05 & 0.01 & 254.73 & 0.05 & 2.96\\
\bottomrule
\end{tabular}
\endgroup{}

\end{table}%

\begin{table}

\caption{\label{tbl-mod-output}Model estimates of proportion of
jellyfish success on index properties and jellyfish hyper-parameters
from the generalised linear model with a binomial family and a logit
link function. The variable smoothness, squintability, number of
jellyfish and maximum number of tries are positively associated with JSO
success rate while data dimension and being flagged as long runtime are
negatively associated with the success rate. The variable squintability
and dimension are significant, suggesting their importance relative to
jellyfish hyper-parameters in the optimisation success.}

\centering{

@@

\begin{tabular}{|>{}lrrr>{}r|}
\toprule
term & estimate & std.error & statistic & p.value\\
\midrule
Intercept & -4.52 & 5.33 & -0.85 & 0.40\\
Smoothness & 1.19 & 1.92 & 0.62 & 0.53\\
Squintability & 2.06 & 0.68 & 3.01 & 0.00\\
Dimension (d) & -0.63 & 0.26 & -2.46 & 0.01\\
Long time & -0.87 & 1.29 & -0.68 & 0.50\\
N. jellyfish & 0.22 & 0.13 & 1.70 & 0.09\\
Max. tries & 0.11 & 0.15 & 0.75 & 0.45\\
\bottomrule
\end{tabular}
\endgroup{}

\end{table}%

\section{Practicalities}\label{sec-discussion}

@@

\section{Conclusion}\label{sec-conclusion}

@@

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-abdulah2023large}
Abdulah, Sameh, Yuxiao Li, Jian Cao, Hatem Ltaief, David Keyes, Marc
Genton, and Ying Sun. 2023. {``Large-Scale Environmental Data Science
with {ExaGeoStatR}.''} \emph{Environmetrics} 34 (1): e2770.
https://doi.org/\url{https://doi.org/10.1002/env.2770}.

\bibitem[\citeproctext]{ref-adams2003sobolev}
Adams, Robert, and John Fournier. 2003. \emph{Sobolev Spaces}. Vol. 140.
Pure and Applied Mathematics. Elsevier.

\bibitem[\citeproctext]{ref-Allaire_Quarto_2022}
Allaire, J. J., Charles Teague, Carlos Scheidegger, Yihui Xie, and
Christophe Dervieux. 2022. \emph{{Quarto}} (version 1.2).
\url{https://doi.org/10.5281/zenodo.5960048}.

\bibitem[\citeproctext]{ref-Bertsimas93}
Bertsimas, Dimitris, and John Tsitsiklis. 1993. {``{Simulated
Annealing}.''} \emph{Statistical Science} 8 (1): 10--15.
\url{https://doi.org/10.1214/ss/1177011077}.

\bibitem[\citeproctext]{ref-chou_novel_2021}
Chou, Jui-Sheng, and Dinh-Nhat Truong. 2021. {``A Novel Metaheuristic
Optimizer Inspired by Behavior of Jellyfish in Ocean.''} \emph{Applied
Mathematics and Computation} 389 (January): 125535.
\url{https://doi.org/10.1016/j.amc.2020.125535}.

\bibitem[\citeproctext]{ref-cook1993projection}
Cook, D., A. Buja, and J. Cabrera. 1993. {``Projection Pursuit Indexes
Based on Orthonormal Function Expansions.''} \emph{Journal of
Computational and Graphical Statistics} 2 (3): 225--50.
\url{https://doi.org/10.2307/1390644}.

\bibitem[\citeproctext]{ref-cook1995grand}
Cook, D., A. Buja, J. Cabrera, and C. Hurley. 1995. {``Grand Tour and
Projection Pursuit.''} \emph{Journal of Computational and Graphical
Statistics} 4 (3): 155--72.
\url{https://doi.org/10.1080/10618600.1995.10474674}.

\bibitem[\citeproctext]{ref-DLMF}
DLMF. 2024. {``{NIST Digital Library of Mathematical Functions}.''}
\url{https://dlmf.nist.gov/10.25}.

>>>>

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{orcidlink}
\usepackage{algorithm}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\theoremstyle{plain}
\newtheorem{assumption}{\protect\assumptionname}
\newtheorem{claim}{\protect\claimname}
\newtheorem{condition}{\protect\conditionname}
\newtheorem{conjecture}{\protect\conjecturename}
\newtheorem{cor}{\protect\corollaryname}
\newtheorem{defn}{\protect\definitionname}
\newtheorem{example}{\protect\examplename}
\newtheorem{lem}{\protect\lemmaname}
\newtheorem{notation}{\protect\notationname}
\newtheorem{problem}{\protect\problemname}
\newtheorem{prop}{\protect\propositionname}
\newtheorem{rem}{\protect\remarkname}
\newtheorem{thm}{\protect\theoremname}
\providecommand{\assumptionname}{Assumption}
\providecommand{\claimname}{Claim}
\providecommand{\conditionname}{Condition}
\providecommand{\conjecturename}{Conjecture}
\providecommand{\corollaryname}{Corollary}
\providecommand{\definitionname}{Definition}
\providecommand{\examplename}{Example}
\providecommand{\lemmaname}{Lemma}
\providecommand{\notationname}{Notation}
\providecommand{\problemname}{Problem}
\providecommand{\propositionname}{Proposition}
\providecommand{\remarkname}{Remark}
\providecommand{\theoremname}{Theorem}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={New Metrics for Assessing Projection Pursuit Indexes, and Guiding Optimisation Choices},
  pdfauthor={H. Sherry Zhang; Dianne Cook; Nicolas Langrené; Jessica Wai Yin Leung},
  pdfkeywords={projection pursuit, jellyfish search optimiser
(JSO), optimisation, grand tour, high-dimensional data, exploratory data
analysis},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@austin.utexas.edu}{\nolinkurl{huize.zhang@austin.utexas.edu}}. Dianne
Cook. Email: \href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}. Nicolas
Langrené. Email: \href{mailto:nicolaslangrene@uic.edu.cn}{\nolinkurl{nicolaslangrene@uic.edu.cn}}. Jessica
Wai Yin
Leung. Email: \href{mailto:Jessica.Leung@monash.edu}{\nolinkurl{Jessica.Leung@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Statistics and Data
Sciences, University of Texas at Austin, Austin, United
States\\ \textsuperscript{2} Department of Econometrics and Business
Statistics, Monash
University, Melbourne, Australia\\ \textsuperscript{3} Department of
Mathematical Sciences, Guangdong Provincial Key Laboratory of
Interdisciplinary Research and Application for Data Science,
BNU-HKBU~United~International~College, Zhuhai, China
\begin{abstract}
The projection pursuit (PP) guided tour interactively optimises a
criteria function known as the PP index, to explore high-dimensional
data by revealing interesting projections. Optimisation of some PP
indexes can be non-trivial, if they are non-smooth functions, or the
optima has a small ``squint angle'', detectable only from close
proximity. To address these challenges, this study investigates the
performance of a recently introduced swarm-based algorithm, Jellyfish
Search Optimiser (JSO), for optimising PP indexes. The performance of
JSO for visualising data is evaluated across various hyper-parameter
settings and compared with existing optimisers. Additionally, methods
for calculating the smoothness and squintability properties of the PP
index are proposed. They are used to assess the optimiser performance in
the presence of PP index complexities. A simulation study illustrates
the use of these performance metrics to compare the JSO with existing
optimisation methods available for the guided tour. The JSO algorithm
has been implemented in the R package, \texttt{tourr}, and functions to
calculate smoothness and squintability are available in the
\texttt{ferrn} package.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
projection pursuit\sep jellyfish search optimiser
(JSO)\sep optimisation\sep grand tour\sep high-dimensional data\sep 
exploratory data analysis
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, interior hidden, breakable, enhanced, frame hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners]}{\end{tcolorbox}}\fi

\setstretch{2}
\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

@@

\hypertarget{sec-PP-properties}{%
\section{Properties of PP indexes}\label{sec-PP-properties}}

@@

\hypertarget{sec-JSO}{%
\section{The jellyfish optimiser}\label{sec-JSO}}

@@

\begin{tcolorbox}[enhanced jigsaw, colbacktitle=quarto-callout-note-color!10!white, breakable, colframe=quarto-callout-note-color-frame, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, leftrule=.75mm, toprule=.15mm, colback=white, bottomrule=.15mm, bottomtitle=1mm, coltitle=black, opacityback=0, title={Algorithm: Jellyfish Optimizer Pseudo Code}, rightrule=.15mm, arc=.35mm, left=2mm]

@@

\hypertarget{tbl-smoothness-squintability}{}
\begin{table}
\caption{\label{tbl-smoothness-squintability}Parameters estimated from the Gaussian process (outputscale \(\eta\),
lengthscale \(\ell\), smoothness \(\nu\), and nugget \(\sigma\)) and
scaled logistic function (\(\theta_1\) to \(\theta_4\) and
\(\varsigma\)) for the pipe-finding and sine-wave finding problems. The
columns \(\nu\) and \(\varsigma\) represent the smoothness and
squintability measures respectively. }\tabularnewline

@@

\begin{tabular}{|>{}lll>{}r|rrr>{}r|rrrr>{}r|}
\toprule
  & shape & index & d & $\eta$ & $\ell$ & $\nu$ & $\sigma$ & $\theta_1$ & $\theta_2$ & $\theta_3$ & $\theta_4$ & $\varsigma$\\
\midrule
1 & pipe & holes & 6 & 0.00 & 0.37 & 2.36 & 0.22 & 1.02 & 0.86 & 3.37 & 0.02 & 3.05\\
2 & pipe & holes & 8 & 0.00 & 0.18 & 2.19 & 0.82 & 1.01 & 0.87 & 3.26 & 0.03 & 2.96\\
3 & pipe & holes & 10 & 0.00 & 0.11 & 2.19 & 3.03 & 1.02 & 0.88 & 3.15 & 0.02 & 2.95\\
4 & pipe & holes & 12 & 0.00 & 0.15 & 2.29 & 1.58 & 1.01 & 0.88 & 3.34 & 0.00 & 3.12\\
5 & sine & MIC & 6 & 0.02 & 0.09 & 2.46 & 0.08 & 0.89 & 0.57 & 1.62 & -0.02 & 1.26\\
6 & sine & MIC & 8 & 0.02 & 0.08 & 2.64 & 0.08 & 0.93 & 0.33 & 1.31 & -0.03 & 0.77\\
7 & sine & TIC & 6 & 0.12 & 0.11 & 2.44 & 0.09 & 0.95 & 0.54 & 1.72 & -0.03 & 1.32\\
8 & sine & TIC & 8 & 0.12 & 0.10 & 2.47 & 0.09 & 0.95 & 0.56 & 1.72 & -0.03 & 1.37\\
9 & sine & dcor2d & 6 & 0.03 & 0.17 & 2.66 & 0.11 & 0.95 & 1.04 & 2.74 & -0.02 & 2.95\\
10 & sine & loess2d & 6 & 0.08 & 0.34 & 1.99 & 0.31 & 1.02 & 1.04 & 2.65 & 0.08 & 2.76\\
11 & sine & splines2d & 6 & 0.04 & 0.24 & 2.54 & 0.10 & 1.01 & 1.05 & 2.73 & -0.01 & 3.12\\
12 & sine & stringy & 6 & 0.00 & 0.01 & 1.54 & 38.17 & 1.05 & 0.01 & 254.73 & 0.05 & 2.96\\
\bottomrule
\end{tabular}
\endgroup{}
\end{table}

\hypertarget{tbl-mod-output}{}
\begin{table}
\caption{\label{tbl-mod-output}Model estimates of proportion of jellyfish success on index properties
and jellyfish hyper-parameters from the generalised linear model with a
binomial family and a logit link function. The variable smoothness,
squintability, number of jellyfish and maximum number of tries are
positively associated with JSO success rate while data dimension and
being flagged as long runtime are negatively associated with the success
rate. The variable squintability and dimension are significant,
suggesting their importance relative to jellyfish hyper-parameters in
the optimisation success. }\tabularnewline

@@

\begin{tabular}{|>{}lrrr>{}r|}
\toprule
term & estimate & std.error & statistic & p.value\\
\midrule
Intercept & -4.52 & 5.33 & -0.85 & 0.40\\
Smoothness & 1.19 & 1.92 & 0.62 & 0.53\\
Squintability & 2.06 & 0.68 & 3.01 & 0.00\\
Dimension (d) & -0.63 & 0.26 & -2.46 & 0.01\\
Long time & -0.87 & 1.29 & -0.68 & 0.50\\
N. jellyfish & 0.22 & 0.13 & 1.70 & 0.09\\
Max. tries & 0.11 & 0.15 & 0.75 & 0.45\\
\bottomrule
\end{tabular}
\endgroup{}
\end{table}

\hypertarget{sec-discussion}{%
\section{Practicalities}\label{sec-discussion}}

@@

\hypertarget{sec-conclusion}{%
\section{Conclusion}\label{sec-conclusion}}

@@

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-abdulah2023large}{}}%
Abdulah, Sameh, Yuxiao Li, Jian Cao, Hatem Ltaief, David Keyes, Marc
Genton, and Ying Sun. 2023. {``Large-Scale Environmental Data Science
with {ExaGeoStatR}.''} \emph{Environmetrics} 34 (1): e2770.
https://doi.org/\url{https://doi.org/10.1002/env.2770}.

\leavevmode\vadjust pre{\hypertarget{ref-adams2003sobolev}{}}%
Adams, Robert, and John Fournier. 2003. \emph{Sobolev Spaces}. Vol. 140.
Pure and Applied Mathematics. Elsevier.

\leavevmode\vadjust pre{\hypertarget{ref-Allaire_Quarto_2022}{}}%
Allaire, J. J., Charles Teague, Carlos Scheidegger, Yihui Xie, and
Christophe Dervieux. 2022. \emph{{Quarto}} (version 1.2).
\url{https://doi.org/10.5281/zenodo.5960048}.

\leavevmode\vadjust pre{\hypertarget{ref-Bertsimas93}{}}%
Bertsimas, Dimitris, and John Tsitsiklis. 1993. {``{Simulated
Annealing}.''} \emph{Statistical Science} 8 (1): 10--15.
\url{https://doi.org/10.1214/ss/1177011077}.

\leavevmode\vadjust pre{\hypertarget{ref-chou_novel_2021}{}}%
Chou, Jui-Sheng, and Dinh-Nhat Truong. 2021. {``A Novel Metaheuristic
Optimizer Inspired by Behavior of Jellyfish in Ocean.''} \emph{Applied
Mathematics and Computation} 389 (January): 125535.
\url{https://doi.org/10.1016/j.amc.2020.125535}.

\leavevmode\vadjust pre{\hypertarget{ref-cook1993projection}{}}%
Cook, D., A. Buja, and J. Cabrera. 1993. {``Projection Pursuit Indexes
Based on Orthonormal Function Expansions.''} \emph{Journal of
Computational and Graphical Statistics} 2 (3): 225--50.
\url{https://doi.org/10.2307/1390644}.

\leavevmode\vadjust pre{\hypertarget{ref-cook1995grand}{}}%
Cook, D., A. Buja, J. Cabrera, and C. Hurley. 1995. {``Grand Tour and
Projection Pursuit.''} \emph{Journal of Computational and Graphical
Statistics} 4 (3): 155--72.
\url{https://doi.org/10.1080/10618600.1995.10474674}.
